Follow-up Samples

### 2017-02-07 submission/manuscript_outline_20161024.Rmd
make intro more about follow ups

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology of the disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an additional increase in the incidence of CRC in the younger population.

This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to recurrence.  Concurrently, there has also been a lot of interest in new areas that could have a role in disease parthenogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the parthenogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.

In this study we investigated what happened to the bacterial microbiome before and after surgery for individuals with either adenoma or carcinoma.  Our analysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.  We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the parthenogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the parthenogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true parthenogenesis of disease.

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].

Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Yet there remains limited information on the bacterial microbiome before and after surgery for removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis of whether that the bacterial microbiome changes after surgery for individuals with a lesion.  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed pre (initial) and post (follow up) surgery.  We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.
